---
title: "Directed Acyclic Graph (DAG) Workflow Demonstration"
author: 
- name: Gavin R Lloyd
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: g.r.lloyd@bham.ac.uk
- name: Ralf J Weber
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: r.j.weber@bham.ac.uk
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2  
    number_sections: true  
    toc_float: true
package: struct
vignette: >
%\VignetteIndexEntry{Directed Acyclic Graph (DAG) Workflow Demonstration}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  ---
  
  ```{r, include=FALSE}
knitr::opts_chunk$set(
dpi=72,fig.width=8,fig.height=6
)
set.seed(57475)
```

# Introduction

This vignette demonstrates how to use the Directed Acyclic Graph (DAG) system in the `struct` package to create and execute complex workflows. We'll use the classic Fisher's Iris dataset to demonstrate a typical data analysis pipeline: mean centering, Principal Component Analysis (PCA), and visualization of the results.

The DAG system allows you to:
- Define complex workflows as directed acyclic graphs
- Execute workflows step-by-step with proper data flow
- Visualize the workflow structure using the `model_dag_chart`
- Handle different types of nodes (data, model, chart, prediction)

# Prerequisites

```{r, eval=FALSE, include=TRUE}
# Install required packages
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Install struct and structToolbox
BiocManager::install(c("struct", "structToolbox"))

# Install DiagrammeR for DAG visualization
if (!requireNamespace("DiagrammeR", quietly = TRUE))
    install.packages("DiagrammeR")
```

```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
    library(struct)
    library(structToolbox)
    library(DiagrammeR)
})
```

# Loading the Data

We'll use the Fisher's Iris dataset, which contains measurements of sepal and petal dimensions for three species of iris flowers.

```{r}
# Load the iris dataset
D = iris_DatasetExperiment()
print(D)
```

The dataset contains 150 samples with 4 features (sepal length, sepal width, petal length, petal width) and a factor indicating the species (setosa, versicolor, virginica).

# Creating the Workflow Components

Before building the DAG, let's create the individual components of our workflow:

## 1. Mean Centering

Mean centering is a preprocessing step that subtracts the mean from each feature, centering the data around zero.

```{r}
# Create mean centering model
mean_centre_model = mean_centre(mode = 'data')
print(mean_centre_model)
```

## 2. Principal Component Analysis (PCA)

PCA is a dimensionality reduction technique that finds the principal components (directions of maximum variance) in the data.

```{r}
# Create PCA model
pca_model = PCA(number_components = 2)
print(pca_model)
```

## 3. PCA Scores Plot

We'll create a chart to visualize the PCA scores, showing how the samples are distributed in the principal component space.

```{r}
# Create PCA scores plot
pca_plot = pca_scores_plot(factor_name = 'Species')
print(pca_plot)
```

# Building the DAG Workflow

Now we'll create a DAG that connects these components in sequence: Data → Mean Centering → PCA → PCA Scores Plot.

## Creating the Nodes

```{r}
# Create data node
data_node = data_node(
    name = "Iris Data",
    description = "Fisher's Iris dataset",
    data = D
)

# Create model nodes
mean_centre_node = model_node(
    name = "Mean Centering",
    description = "Center the data by subtracting column means",
    model = mean_centre_model
)

pca_node = model_node(
    name = "PCA",
    description = "Principal Component Analysis with 2 components",
    model = pca_model
)

# Create chart node
pca_chart_node = chart_node(
    name = "PCA Scores Plot",
    description = "Visualize PCA scores colored by species",
    chart = pca_plot)

autoscale_node = model_node(
    name='Autoscale',
    description='Autoscale',
    model=autoscale())

pls_node = model_node(
    mode = PLSDA(factor_name = 'Species')
)

pls_chart_node = chart_node(
    chart = pls_scores_plot(factor_name = 'chart')
)

scree_node = chart_node(
    chart = pca_scree_plot()
)
```

## Defining the Edges

The edges define the flow of data between nodes:

```{r}
# Define edges (connections between nodes)
edges = list(
    # Data flows to mean centering
    list(from = "Iris Data", from_param = "asis", 
         to = "Mean Centering", to_param = "input_data"),
    
    # Mean centered data flows to PCA
    list(from = "Mean Centering", from_param = "predicted", 
         to = "PCA", to_param = "input_data"),
    
    # PCA model flows to chart
    list(from = "PCA", from_param = "asis", 
         to = "PCA Scores Plot", to_param = "input_object"),
    
    list(from = "PCA", from_param = "asis", 
         to = "Scree Plot", to_param = "input_object"),
    
    list(from = "Mean Centering", from_param = "predicted", 
         to = "PLS", to_param = "input_object"),
    
    # PCA model flows to chart
    list(from = "PLS", from_param = "asis", 
         to = "PLS Scores Plot", to_param = "input_object")
)
```

## Creating the DAG

```{r}
# Create the DAG
iris_dag = model_dag(
    name = "Iris Analysis Workflow",
    description = "Mean centering followed by PCA and visualization",
    edges = edges,
    nodes = list(
        "Iris Data" = data_node,
        "Mean Centering" = mean_centre_node,
        "PCA" = pca_node,
        "PCA Scores Plot" = pca_chart_node,
        "PLS" = pls_node,
        "PLS Scores Plot" = pls_chart_node,
        "Scree Plot" = scree_node
    )
)

print(iris_dag)
```

# Visualizing the DAG Structure

We can visualize the DAG structure using the `model_dag_chart`:

```{r}
# Create DAG chart
dag_chart = model_dag_chart(
    name = "Iris Workflow Chart",
    description = "Visualization of the iris analysis workflow",
    node_width = 'auto',
    node_height = 'auto',
    font_size = 6,
    show_params = 'changed',
    arrow_size = 0.5,
    arrow_shape = 'normal',
    max_param_length = Inf,
    layout='tree'
)

# Plot the DAG
chart_plot(dag_chart, iris_dag)
```

The visualization shows:
- **Iris Data**: The input dataset
- **Mean Centering**: The preprocessing step with its parameters
- **PCA**: The dimensionality reduction step with number of components
- **PCA Scores Plot**: The visualization step

The arrows show the data flow between the nodes.

# Executing the DAG

Now let's execute the DAG to perform the analysis:

```{r}
# Execute the DAG
results = dag_execute(iris_dag,verbose=FALSE)
```

The execution process:
1. **Data Node**: Provides the iris dataset
2. **Mean Centering**: Centers the data by subtracting column means
3. **PCA**: Performs principal component analysis on the centered data
4. **Chart Node**: Creates the PCA scores plot

# Examining the Results

Let's examine what was produced by each step:

## Mean Centering Results

```{r}
# Get the mean centered data
mean_centered_data = results$results$`Mean Centering`
print(mean_centered_data)
```

## PCA Results

```{r}
# Get the PCA results
pca_results = results$results$`PCA`
print(pca_results)
```

## PCA Scores Plot

```{r}
# Get the PCA scores plot
pca_plot_result = results$results$`PCA Scores Plot`
print(pca_plot_result)
```

# Understanding the Data Flow

The DAG system ensures that data flows correctly between steps:

1. **Raw Data** → **Mean Centering**: The original iris data is passed to the mean centering step
2. **Mean Centered Data** → **PCA**: The centered data is used as input for PCA
3. **PCA Model** → **Chart**: The trained PCA model is passed to the plotting function
4. **Mean Centered Data** → **Chart**: The centered data is also passed to the chart for visualization

This ensures that:
- The PCA is performed on properly preprocessed data
- The visualization uses the correct data and model
- All steps are executed in the correct order

# Customizing the Workflow

You can easily modify the workflow by changing parameters or adding new steps:

## Example: Adding More Components

```{r}
# Create a more complex workflow with additional preprocessing
# Add log transformation
log_transform_model = log_transform()
log_node = model_node(
    name = "Log Transform",
    description = "Log transform the data",
    model = log_transform_model
)

# Create a new DAG with log transformation
complex_edges = list(
    list(from = "Iris Data", from_param = "asis", 
         to = "Log Transform", to_param = "input_data"),
    list(from = "Log Transform", from_param = "predicted", 
         to = "Mean Centering", to_param = "input_data"),
    list(from = "Mean Centering", from_param = "predicted", 
         to = "PCA", to_param = "input_data"),
    list(from = "PCA", from_param = "asis", 
         to = "PCA Scores Plot", to_param = "input_object")
)

complex_dag = model_dag(
    name = "Complex Iris Analysis",
    description = "Log transform + mean centering + PCA",
    edges = complex_edges,
    nodes = list(
        "Iris Data" = data_node,
        "Log Transform" = log_node,
        "Mean Centering" = mean_centre_node,
        "PCA" = pca_node,
        "PCA Scores Plot" = pca_chart_node
    )
)

# Visualize the complex workflow
chart_plot(dag_chart, complex_dag)
```

## Example: Different PCA Parameters

```{r}
# Create PCA with different number of components
pca_3comp = PCA(number_components = 3)
pca_3comp_node = model_node(
    name = "PCA (3 components)",
    description = "PCA with 3 principal components",
    model = pca_3comp
)

# Create a DAG with 3-component PCA
pca3_edges = list(
    list(from = "Iris Data", from_param = "asis", 
         to = "Mean Centering", to_param = "input_data"),
    list(from = "Mean Centering", from_param = "predicted", 
         to = "PCA (3 components)", to_param = "input_data")
)

pca3_dag = model_dag(
    name = "Iris Analysis with 3 PCA Components",
    description = "Mean centering followed by 3-component PCA",
    edges = pca3_edges,
    nodes = list(
        "Iris Data" = data_node,
        "Mean Centering" = mean_centre_node,
        "PCA (3 components)" = pca_3comp_node
    )
)

# Visualize
chart_plot(dag_chart, pca3_dag)
```

# Advanced DAG Features

## Conditional Execution

You can create more complex workflows with conditional execution:

```{r}
# Example: Create a workflow that can handle different preprocessing options
# This is a conceptual example showing how you might structure conditional workflows

# Create multiple preprocessing options
autoscale_model = autoscale()
pareto_model = pareto_scale()

# Create nodes for different preprocessing methods
autoscale_node = model_node(
    name = "Autoscaling",
    description = "Autoscale the data (mean=0, std=1)",
    model = autoscale_model
)

pareto_node = model_node(
    name = "Pareto Scaling",
    description = "Pareto scale the data",
    model = pareto_model
)

# Note: In practice, you would use conditional logic or separate DAGs
# for different preprocessing strategies
```

## Error Handling

The DAG system includes error handling to ensure robust execution:

```{r}
# Example: What happens if a step fails
# The DAG execution will stop and provide informative error messages
# This helps identify and fix issues in the workflow

# Try executing with invalid parameters
tryCatch({
    invalid_pca = PCA(number_components = 10)  # More components than features
    invalid_node = model_node(
        name = "Invalid PCA",
        description = "PCA with too many components",
        model = invalid_pca
    )
    
    invalid_edges = list(
        list(from = "Iris Data", from_param = "asis", 
             to = "Invalid PCA", to_param = "input_data")
    )
    
    invalid_dag = model_dag(
        name = "Invalid Workflow",
        description = "Workflow with invalid parameters",
        edges = invalid_edges,
        nodes = list(
            "Iris Data" = data_node,
            "Invalid PCA" = invalid_node
        )
    )
    
    # This would fail during execution
    # dag_execute(invalid_dag)
    
}, error = function(e) {
    cat("Error caught:", e$message, "\n")
})
```

# Best Practices for DAG Workflows

## 1. Clear Naming

Use descriptive names for nodes and parameters:

```{r}
# Good naming
good_node = model_node(
    name = "PCA with 2 Components",
    description = "Principal Component Analysis retaining 2 components",
    model = pca_model
)

# Avoid vague names
# bad_node = model_node(name = "Model1", ...)
```

## 2. Proper Documentation

Include descriptions for all nodes and the overall workflow:

```{r}
# Well-documented DAG
well_documented_dag = model_dag(
    name = "Iris Species Classification Pipeline",
    description = "Complete workflow for iris species classification including preprocessing, dimensionality reduction, and visualization",
    edges = edges,
    nodes = list(...)
)
```

## 3. Modular Design

Break complex workflows into smaller, reusable components:

```{r}
# Create reusable preprocessing module
preprocessing_dag = model_dag(
    name = "Standard Preprocessing",
    description = "Common preprocessing steps: mean centering and scaling",
    edges = list(
        list(from = "Input Data", from_param = "asis", 
             to = "Mean Centering", to_param = "input_data"),
        list(from = "Mean Centering", from_param = "asis", 
             to = "Autoscaling", to_param = "input_data")
    ),
    nodes = list(
        "Input Data" = data_node,
        "Mean Centering" = mean_centre_node,
        "Autoscaling" = autoscale_node
    )
)
```

## 4. Validation

Always validate your DAG before execution:

```{r}
# Check that all nodes are properly connected
# Verify that required parameters are set
# Ensure the workflow makes logical sense

# Example validation checks
cat("Number of nodes:", length(iris_dag$nodes), "\n")
cat("Number of edges:", length(iris_dag$edges), "\n")
cat("All nodes have names:", all(!is.null(names(iris_dag$nodes))), "\n")
```

# Conclusion

The DAG system in the `struct` package provides a powerful and flexible way to create complex data analysis workflows. Key benefits include:

- **Reproducibility**: Workflows are explicitly defined and can be re-executed
- **Modularity**: Components can be reused and combined in different ways
- **Visualization**: The workflow structure can be visualized using `model_dag_chart`
- **Error Handling**: Robust execution with informative error messages
- **Flexibility**: Easy to modify and extend workflows

The iris dataset example demonstrates a typical workflow, but the same principles apply to any data analysis pipeline. The DAG system is particularly useful for:

- Metabolomics data analysis
- Machine learning pipelines
- Quality control workflows
- Multi-step statistical analyses

By using the DAG system, you can create complex, reproducible workflows that are easy to understand, modify, and share with others.

# Session Information

```{r}
sessionInfo()
``` 
