---
title: "Directed Acyclic Graphs (DAGs) for struct Models"
author: 
- name: Gavin R Lloyd
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: g.r.lloyd@bham.ac.uk
- name: Ralf J Weber
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: r.j.weber@bham.ac.uk
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2  
    number_sections: true  
    toc_float: true
package: struct
vignette: >
  %\VignetteIndexEntry{Directed Acyclic Graphs (DAGs) for struct Models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    dpi=72,fig.width=5,fig.height=5.5
)
set.seed(57475)
```

# Introduction

This vignette demonstrates the Directed Acyclic Graph (DAG) functionality in the `struct` package. DAGs provide a flexible way to create complex workflows by connecting nodes that represent data, models, and operations.

## Key Concepts

The DAG functionality consists of five main classes:

1. **`model_dag`** - Represents a directed acyclic graph with edges defining the workflow
2. **`struct_node`** - Base class for all node types (data_node, model_node, prediction_node)
3. **`model_node`** - Represents a node containing a struct model (uses `model_apply`)
4. **`prediction_node`** - Represents a node that receives a trained model via an edge and uses `model_predict`
5. **`data_node`** - Represents a node containing a DatasetExperiment object

## Getting Started

```{r, message=FALSE,warning=FALSE}
suppressPackageStartupMessages({
  # load the package
  library(struct)
  
  # load ggplot for plotting
  library(ggplot2)
    
    library(structToolbox)
})
```

# Edge Specification System

The DAG system uses a flexible edge specification that allows precise control over how data flows between nodes. Each edge specifies:

- **`from`**: Source node name
- **`from_param`**: Parameter/slot to get from source node
- **`to`**: Target node name  
- **`to_param`**: Parameter/slot to set in target node

## Parameter Types

### from_param values:
- **`"asis"`**: Pass the object without change (default for data nodes)
- **`"predicted"`**: Use `predicted()` output from model nodes
- **Named parameter**: Access specific slots/parameters from struct objects

### to_param values:
- **`"input_data"`**: Set the input_data slot (default for model nodes)
- **`"input_model"`**: Set the input_model slot (for prediction nodes)
- **Named parameter**: Set a specific parameter on the target model

## Default Behavior

- **Data nodes**: `from_param = "asis"` (pass DatasetExperiment as-is)
- **Model nodes**: `from_param = "predicted"` (use predicted output), `to_param = "input_data"`
- **Prediction nodes**: `from_param = "asis"` (pass model as-is), `to_param = "input_model"` or `"input_data"`

# Simple DAG Workflow

Let's start with a simple preprocessing workflow: Data → Mean Centering → PCA

```{r}
# Load example data
D = iris_DatasetExperiment()

# Create a simple preprocessing workflow:
# Data -> Mean Center -> PCA

# Step 1: Create data node
data_node1 = data_node(
    name = 'Raw Data',
    description = 'Iris dataset',
    data = D
)

# Step 2: Create mean centering node
mean_center_model = mean_centre()
mean_center_node = model_node(
    name = 'Mean Centering',
    description = 'Center the data by subtracting the mean',
    model = mean_center_model
)

# Step 3: Create PCA node
pca_model = PCA()
pca_node = model_node(
    name = 'PCA',
    description = 'Principal Component Analysis',
    model = pca_model
)

# Create the DAG with nodes included
dag = model_dag(
    name = 'Preprocessing Workflow',
    description = 'A simple preprocessing workflow: Mean centering followed by PCA',
    edges = list(
        list(from = 'Raw Data', from_param = 'asis', to = 'Mean Centering', to_param = 'input_data'),
        list(from = 'Mean Centering', from_param = 'predicted', to = 'PCA', to_param = 'input_data')
    ),
    nodes = list(
        'Raw Data' = data_node1,
        'Mean Centering' = mean_center_node,
        'PCA' = pca_node
    )
)

# Execute DAG (much simpler!)
dag = dag_execute(dag, verbose = FALSE)

# Create a PCA scores plot
C = pca_scores_plot(factor_name = 'Species')
chart_plot(C, dag$results$PCA)

C = model_dag_chart()
chart_plot(C, dag)
```


# Complex DAG with Parallel Paths

This example demonstrates a more complex workflow with parallel paths and different model modes:

```{r}
# Create a more complex workflow:
# Data -> [Mean Center, autoscale] -> [PCA, PLS] -> Results

# Step 1: Create data node
input_node = data_node(
    name = 'Raw Data',
    description = 'Iris dataset',
    data = iris_DatasetExperiment()
)

# Step 2: Create mean centering node
mean_center_node = model_node(
    name = 'Mean Centering',
    description = 'Center the data by subtracting the mean',
    model = mean_centre()
)

autoscale_node = model_node(
    name = 'Autoscaling',
    description = 'Center the data by subtracting the mean and then divide by the standard deviation',
    model = autoscale()
)

# Step 3: Create PCA node
pca_node = model_node(
    name = 'PCA',
    description = 'PCA',
    model = PCA()
)

# Step 5: Create PLS node
pls_node = model_node(
    name = 'PLS Train',
    description = 'Train PLS model',
    model = PLSDA(factor_name='Species')
)

# Create the DAG with parallel paths and nodes included
dag = model_dag(
    name = 'Complex Analysis Workflow',
    description = 'A complex workflow with parallel PCA and PLS analysis',
    edges = list(
        list(from = 'Raw Data', from_param = 'asis', to = 'Mean Centering', to_param = 'input_data'),
        list(from = 'Mean Centering', from_param = 'predicted', to = 'PCA', to_param = 'input_data'),
        list(from = 'Raw Data', from_param = 'asis', to = 'Autoscale', to_param = 'input_data'),
        list(from = 'Autoscale', from_param = 'predicted', to = 'PLS', to_param = 'input_data')
    ),
    nodes = list(
        'Raw Data' = input_node,
        'Mean Centering' = mean_center_node,
        'PCA' = pca_node,
        'PLS' = pls_node,
        'Autoscale' = autoscale_node
    )
)

# Execute the DAG
cat('Executing complex DAG workflow...\n')
dag = dag_execute(dag, verbose = FALSE)

C = model_dag_chart(box_width=0.5)
chart_plot(C, dag)
```

```{r}
# Create a PCA scores plot
C = pca_scores_plot(factor_name = 'Species')
g1=chart_plot(C, dag$results$PCA)

# Create a PLS scores plot
C = plsda_scores_plot(factor_name = 'Species')
g2=chart_plot(C, dag$results$PLS)

cowplot::plot_grid(g1,g2)
```

# Model Apply Emulation

This example demonstrates how to emulate `model_apply` using separate `model_train` and `model_predict` nodes:

```{r}
# Load example data
D = iris_DatasetExperiment()

# Create a workflow that emulates model_apply using separate train and predict steps:
# Data -> PCA Train -> PCA Predict (equivalent to model_apply)

# Step 1: Create data node
data_node1 = data_node(
    name = 'Raw Data',
    description = 'Iris dataset',
    data = D
)

# Step 2: Create PCA train node
pca_model_train = PCA()
pca_train_node = model_node(
    name = 'PCA Train',
    description = 'Train PCA model (first step of model_apply)',
    model = pca_model_train
)

# Step 3: Create PCA predict node (model will be provided via edge)
pca_predict_node = prediction_node(
    name = 'PCA Predict',
    description = 'Apply trained PCA model (second step of model_apply)'
)

# Create the DAG that emulates model_apply with nodes included
dag = model_dag(
    name = 'Model Apply Emulation',
    description = 'Emulating model_apply using separate train and predict nodes',
    edges = list(
        list(from = 'Raw Data', from_param = 'asis', to = 'PCA Train', to_param = 'input_data'),
        list(from = 'Raw Data', from_param = 'asis', to = 'PCA Predict', to_param = 'input_data'),  # Data input
        list(from = 'PCA Train', from_param = 'asis', to = 'PCA Predict', to_param = 'input_model')  # Trained model input
    ),
    nodes = list(
        'Raw Data' = data_node1,
        'PCA Train' = pca_train_node,
        'PCA Predict' = pca_predict_node
    )
)

# Execute the DAG
cat('Executing model_apply emulation DAG...\n')
cat('This demonstrates how model_apply = model_train + model_predict\n')
dag = dag_execute(dag, verbose = TRUE)

C = model_dag_chart()
chart_plot(C, dag)
```

# Node Types and Execution Logic

The DAG system uses different node types with specific execution logic and input_data slots:

## Node Types

All node types inherit from `struct_node` and include an `input_data` slot:

- **`data_node`** - Contains a DatasetExperiment object with `input_data` slot
- **`model_node`** - Contains a struct model with `input_data` slot (uses `model_apply`)
- **`prediction_node`** - Has `input_data` and `input_model` slots (uses `model_predict`)

## Execution Logic

1. **Data nodes**: Pass through their DatasetExperiment data to connected nodes
2. **Model nodes**: Use `model_apply` with input data from the `input_data` slot
3. **Prediction nodes**: Use `model_predict` with trained models and input data
   - **Requires two inputs**: One edge providing a trained model to `input_model`, one edge providing input data to `input_data`

## Input Data Flow

- **Data → Model**: DatasetExperiment passed to model's `input_data` slot
- **Model → Model**: Predicted output passed to next model's `input_data` slot  
- **Model → Prediction**: Trained model passed to prediction node's `input_model` slot
- **Data → Prediction**: DatasetExperiment passed to prediction node's `input_data` slot

# Benefits of DAGs

The DAG functionality provides several benefits:

1. **Self-contained Workflows** - DAG objects contain structure, nodes, and results
2. **Modularity** - Each step is encapsulated in its own node
3. **Reusability** - Nodes can be reused in different workflows
4. **Clarity** - The workflow structure is explicitly defined
5. **Validation** - Automatic validation of workflow structure
6. **Flexibility** - Support for complex workflows with parallel paths
7. **Simplicity** - Clear separation between training (`model_node`) and prediction (`prediction_node`) operations
8. **Automatic Data Flow** - Intelligent handling of data flow between nodes
9. **Object-oriented Design** - `dag_execute` is a method of the DAG class
10. **Integrated Results** - Results are stored within the DAG object for easy access

# Conclusion

Directed Acyclic Graphs provide a powerful and flexible way to create complex workflows in the `struct` package. They allow you to:

- Create self-contained workflow objects that contain structure, nodes, and results
- Break down complex operations into their constituent parts
- Create parallel processing paths
- Inspect intermediate model states
- Reuse components across different workflows
- Maintain clear workflow documentation
- Automatically handle train/predict workflows with dedicated node types
- Access results directly from the DAG object

The DAG approach complements the existing `model_seq` functionality while providing additional flexibility for complex workflows. The object-oriented design with integrated nodes and results makes DAGs easy to create, execute, and analyze. The simplified node types (`model_node` for training, `prediction_node` for prediction) make it easy to create and understand workflows. 
