---
title: "Directed Acyclic Graphs (DAGs) for struct Models"
author: 
- name: Gavin R Lloyd
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: g.r.lloyd@bham.ac.uk
- name: Ralf J Weber
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: r.j.weber@bham.ac.uk
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2  
    number_sections: true  
    toc_float: true
package: struct
vignette: >
  %\VignetteIndexEntry{Directed Acyclic Graphs (DAGs) for struct Models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    dpi=72,fig.width=5,fig.height=5.5
)
set.seed(57475)
```

# Introduction

This vignette demonstrates the Directed Acyclic Graph (DAG) functionality in the `struct` package. DAGs provide a flexible way to create complex workflows by connecting nodes that represent data, models, and operations.

## Key Concepts

The DAG functionality consists of five main classes:

1. **`model_dag`** - Represents a directed acyclic graph with edges defining the workflow
2. **`struct_node`** - Base class for all node types (data_node, model_node, prediction_node)
3. **`model_node`** - Represents a node containing a struct model (uses `model_apply`)
4. **`prediction_node`** - Represents a node that receives a trained model via an edge and uses `model_predict`
5. **`data_node`** - Represents a node containing a DatasetExperiment object

## Getting Started

```{r, message=FALSE,warning=FALSE}
suppressPackageStartupMessages({
  # load the package
  library(struct)
  
  # load ggplot for plotting
  library(ggplot2)
})
```

# Simple DAG Workflow

Let's start with a simple preprocessing workflow: Data → Mean Centering → PCA

```{r}
# Load example data
D = iris_DatasetExperiment()

# Create a simple preprocessing workflow:
# Data -> Mean Center -> PCA

# Step 1: Create data node
data_node1 = data_node(
    name = 'Raw Data',
    description = 'Iris dataset',
    data = D
)

# Step 2: Create mean centering node
mean_center_model = mean_centre()
mean_center_node = model_node(
    name = 'Mean Centering',
    description = 'Center the data by subtracting the mean',
    model = mean_center_model
)

# Step 3: Create PCA node
pca_model = PCA()
pca_node = model_node(
    name = 'PCA',
    description = 'Principal Component Analysis',
    model = pca_model
)

# Create the DAG
dag = model_dag(
    name = 'Preprocessing Workflow',
    description = 'A simple preprocessing workflow: Mean centering followed by PCA',
    edges = list(
        list(from = 'Raw Data', to = 'Mean Centering'),
        list(from = 'Mean Centering', to = 'PCA')
    )
)

# Create named list of nodes
nodes = list(
    'Raw Data' = data_node1,
    'Mean Centering' = mean_center_node,
    'PCA' = pca_node
)

# Execute the DAG
cat('Executing DAG workflow...\n')
results = dag_execute(dag, nodes, verbose = TRUE)
```

Now let's visualize the PCA results:

```{r}
# Create a PCA scores plot
C = pca_scores_plot(factor_name = 'Species')
chart_plot(C, results$PCA)
```

# Complex DAG with Parallel Paths

This example demonstrates a more complex workflow with parallel paths and different model modes:

```{r}
# Create a more complex workflow:
# Data -> [Mean Center, autoscale] -> [PCA, PLS] -> Results

# Step 1: Create data node
input_node = data_node(
    name = 'Raw Data',
    description = 'Iris dataset',
    data = iris_DatasetExperiment()
)

# Step 2: Create mean centering node
mean_center_node = model_node(
    name = 'Mean Centering',
    description = 'Center the data by subtracting the mean',
    model = mean_centre()
)

autoscale_node = model_node(
    name = 'Autoscaling',
    description = 'Center the data by subtracting the mean and then divide by the standard deviation',
    model = autoscale()
)

# Step 3: Create PCA node
pca_node = model_node(
    name = 'PCA',
    description = 'PCA',
    model = PCA()
)

# Step 5: Create PLS node
pls_node = model_node(
    name = 'PLS Train',
    description = 'Train PLS model',
    model = PLSDA(factor_name='Species')
)

# Create the DAG with parallel paths
dag = model_dag(
    name = 'Complex Analysis Workflow',
    description = 'A complex workflow with parallel PCA and PLS analysis',
    edges = list(
        list(from = 'Raw Data', to = 'Mean Centering'),
        list(from = 'Mean Centering', to = 'PCA'),
        list(from = 'Raw Data', to = 'Autoscale'),
        list(from = 'Autoscale', to = 'PLS')
    )
)

# Create named list of nodes
nodes = list(
    'Raw Data' = input_node,
    'Mean Centering' = mean_center_node,
    'PCA' = pca_node,
    'PLS' = pls_node,
    'Autoscale' = autoscale_node
)

# Execute the DAG
cat('Executing complex DAG workflow...\n')
results = dag_execute(dag, nodes, verbose = TRUE)
```

```{r}
# Create a PCA scores plot
C = pca_scores_plot(factor_name = 'Species')
g1=chart_plot(C, results$PCA)

# Create a PLS scores plot
C = plsda_scores_plot(factor_name = 'Species')
g2=chart_plot(C, results$PLS)

cowplot::plot_grid(g1,g2)
```

# Model Apply Emulation

This example demonstrates how to emulate `model_apply` using separate `model_train` and `model_predict` nodes:

```{r}
# Load example data
D = iris_DatasetExperiment()

# Create a workflow that emulates model_apply using separate train and predict steps:
# Data -> PCA Train -> PCA Predict (equivalent to model_apply)

# Step 1: Create data node
data_node1 = data_node(
    name = 'Raw Data',
    description = 'Iris dataset',
    data = D
)

# Step 2: Create PCA train node
pca_model_train = PCA()
pca_train_node = model_node(
    name = 'PCA Train',
    description = 'Train PCA model (first step of model_apply)',
    model = pca_model_train
)

# Step 3: Create PCA predict node (model will be provided via edge)
pca_predict_node = prediction_node(
    name = 'PCA Predict',
    description = 'Apply trained PCA model (second step of model_apply)'
)

# Create the DAG that emulates model_apply
dag = model_dag(
    name = 'Model Apply Emulation',
    description = 'Emulating model_apply using separate train and predict nodes',
    edges = list(
        list(from = 'Raw Data', to = 'PCA Train'),
        list(from = 'Raw Data', to = 'PCA Predict'),  # Data input
        list(from = 'PCA Train', to = 'PCA Predict')  # Trained model input
    )
)

# Create named list of nodes
nodes = list(
    'Raw Data' = data_node1,
    'PCA Train' = pca_train_node,
    'PCA Predict' = pca_predict_node
)

# Execute the DAG
cat('Executing model_apply emulation DAG...\n')
cat('This demonstrates how model_apply = model_train + model_predict\n')
results = dag_execute(dag, nodes, verbose = TRUE)
```

# Node Types and Execution Logic

The DAG system uses different node types with specific execution logic:

## Node Types

All node types inherit from `struct_node`:

- **`data_node`** - Contains a DatasetExperiment object
- **`model_node`** - Contains a struct model (uses `model_apply`)
- **`prediction_node`** - Receives a trained model via an edge and uses `model_predict`

## Execution Logic

1. **Data nodes**: Pass through their DatasetExperiment data to connected nodes
2. **Model nodes**: Use `model_apply` with input data (DatasetExperiment or predicted output from previous models)
3. **Prediction nodes**: Use `model_predict` with trained models and input data
   - **Requires two inputs**: One edge providing a trained model, one edge providing input data

# Benefits of DAGs

The DAG functionality provides several benefits:

1. **Modularity** - Each step is encapsulated in its own node
2. **Reusability** - Nodes can be reused in different workflows
3. **Clarity** - The workflow structure is explicitly defined
4. **Validation** - Automatic validation of workflow structure
5. **Flexibility** - Support for complex workflows with parallel paths
6. **Simplicity** - Clear separation between training (`model_node`) and prediction (`prediction_node`) operations
7. **Automatic Data Flow** - Intelligent handling of data flow between nodes

# Conclusion

Directed Acyclic Graphs provide a powerful and flexible way to create complex workflows in the `struct` package. They allow you to:

- Break down complex operations into their constituent parts
- Create parallel processing paths
- Inspect intermediate model states
- Reuse components across different workflows
- Maintain clear workflow documentation
- Automatically handle train/predict workflows with dedicated node types

The DAG approach complements the existing `model_seq` functionality while providing additional flexibility for complex workflows. The simplified node types (`model_node` for training, `prediction_node` for prediction) make it easy to create and understand workflows. 
